An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
Continue fixing red

\subsubsection{Example~1: String length}
\Fudge{We have already illustrated refer intro}
A useful example of such a simplifying morphism can express string length, the
problem that originally motivated our study of the Parikh map. We let $\Monoid =
\left(\mathbb{Z};+;0\right)$ and let $\Map(x) = 1$. Then the length of a string
$s = s_1 \RepeatSum{\Concat}  s_n$ is given by $\Map(s) = \sum \Map(s_i) = 1
\RepeatSum{+} 1$. Note that if we substitute vectors and vector addition for
this monoid, we obtain the regular Parikh map.

Intuitively, this produces a simpler problem than computing the Parikh image, as
we now ignore not just the order of characters but also their differences. This
means that the branching automaton of \cref{fig:len-branch} becomes
equivalent to the non-branching automaton of \cref{fig:len-branch-free}.
Note that the elimination of the choice of paths also effectively eliminates the
possibility of disconnecting one of the self-loops. This, in turn, means that we
can eliminate all of the connectedness implications of the quantified formula,
which would in turn simplify it (and the quantifier elimination of it)
considerably.

\subsubsection{Example~2: Parikh Automata}\label{sec:parikh-automata}
\contents{\item Ostrich+!}

@@

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where $C \subseteq \Naturals^d$ is a semi-linear set (or, equivalently, a Presburger formula), where $\Automaton$ is a finite automaton with the alphabet $\Alphabet \times D$, where $D \subseteq \Naturals^d$.

  We say that $\Tuple{\Automaton, \varphi}$ \emph{recognises} the language, $\Language(\Automaton, C)$, that is it recognises the language crated by is the $\pi_\Alphabet$-projection on $\Alphabet$ whose $\pi_{v} \subseteq C$.
\end{definition}

\begin{figure}[ht]
  \centering
      \includegraphics[width=0.75\linewidth]{parikh_automaton}
      \caption{A The automaton part of a Parikh automaton for $\AcaOrBc{}$ with
      $\Alphabet = \Set{\text{a, b, c}}, d = 3$. The semilinear set/Presburger
      formula containing the constraints on the final register values cannot be
      visualised.}\label{fig:parikh-automaton}
    \end{figure}

@@

We assume an NFA $\Automaton = \AutomatonTuple$ with $\NrTransitions$
transitions $\Transitions =
\Set{\EllipsisSequence{\Transitions}{\NrTransitions}}$. For convenience, we then introduce the following supporting notations:

\begin{definition}
  A \textit{path} $\Path = \PathEnumeration$ of an automaton~$\Automaton$ with
  transitions $\Transitions$ represents a path through $\Automaton$ using
  transitions in $\Transitions$ (i.e. $\Transitions(\State_k, \Label_{k+1},
  \State_{k+1})$ holds for every $k$), passing zero or more labels $\Label_1,
  \ldots, \Label_n$. The path must begin in the initial state, i.e.~$\State_0 =
  \InitialState$. However, the ending state, $\State_n$, is not necessarily an
  accepting state.
  \end{definition}

\begin{definition}
  Moreover, we also talk about the \textit{set of paths} of an automaton,
  $\Paths(\Automaton)$, a possibly infinite (if $\Automaton$ has loops) set of
  valid paths through $\Automaton$. Additionally, we use the
  notation~$\Paths(\Automaton, \State)$ to mean all paths ending in
  state~$\State$.
\end{definition}

\begin{definition}
  The \textit{word} of a path $\WordOf(\Path) = \Label_1 \RepeatSum{\Concat} \Label_n$ is
  the word read out on its labels.
\end{definition}

\begin{definition}
  \Fudge{fixme: use pattern matching on path definition}
  The \textit{states} of a path $\StatesOf(\Path) = \Set{\State \SuchThat
  \FromLabelTo{\State}{}{\State'} \in \Path \text{ or }
  \FromLabelTo{\State'}{}{\State} \in \Path}$ are the states visited along a
  path. Note that $\InitialState \in \StatesOf(\Path)$ for every path since all
  paths start in the initial state.
\end{definition}

\begin{definition}
  \Fudge{if every valid path containing q overlaps with C}
  A \textit{cut}, $C$ of an automaton~$\Automaton = \AutomatonTuple$,
  $\SeparatingCut(C, \Automaton, \State)$, is a minimal set of transitions whose
  removal from~$\Transitions$ would cause~$\State \in \States$ to be unreachable
  from the initial state~$\InitialState$, or cut it off from reaching any
  accepting state. $\SeparatingCut(\Automaton, \InitialState)$ is crucially
  undefined if $\Automaton$ accepts any string since it cannot be made unreachable
  from itself by definition and must have at least onepath to an accepting state
  if it accepts any string.
\end{definition}

@@

Finally, we present the rules of our calculus for one automaton in
\cref{tbl:rules:single}. Note that we operate on sets of symbols (terms,
clauses). Additionally, we also use the convention of splitting the formulas
into linear inequalities ($\SomeInequalities$) and any other clauses
($\SomeClause$).

We use the shorthand notation~$\Transitions_\Automaton$ to refer to the
transitions of an automaton~$\Automaton$. Additionally, for an
automaton~$\Automaton = \AutomatonTuple$ we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$.

These rules are meant to be executed in an automated theorem prover, and are
read bottom-up. Since a large part of our rules operate by adding and literally
matching linear (in)equalities in a proof goal, we use the shorthand of listing
the matched inequalities as antecedents. An example of this can be seen in
the~\Propagate{} rule.

Please note that the filtering function~$\Filter$ is evaluated symbolically in
these rules, and can in practice be read as a symbolic function from transitions
to terms (e.g.~\texttt{t} or~\texttt{t+1}), not dissimilar to~$\FlowEq$ below.
In our implementation, $\Filter$~is a vector of fresh unknown constants with the
same size as~$\Transitions$.

We assume that every rule only fires when it would add a new clause. For
example, this means that we cannot use~\Split{} to split on the same term twice.
This suggests a proof strategy where you \Propagate{} when you can, \Split{}
when you must, and \Subsume{} when neither is possible anymore.

  % EXPAND
  \Expand & 
    \inferrule
  {\Connected(\Automaton, \Filter) \land \FlowEq(\Automaton, \Filter) \land \MonoidElement = \sum_{\Transition \in \Transitions_\Automaton} \Filter(\Transition) \cdot \Map(\Transition), \SomeInequalities, \SomeClause}
  {\SinglePredicateInstance, \SomeInequalities, \SomeClause} & 
  None \\[5ex]

  % SPLIT
  \Split & 
  \inferrule{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) = 0 \mid \Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) > 0}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  if $\Transition \in \Transitions_\Automaton$ \\[5ex]

  % PROPAGATE
  \Propagate &
  \inferrule{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause, \Filter(\Transition) = 0}{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause} &
  if $t = \FromLabelTo{\State}{}{\State'} \in \Transitions_\Automaton, \SeparatingCut(C, \Automaton, \State)$ \\[5ex]

  % SUBSUME
  \Subsume &
  \inferrule{\SomeInequalities,\SomeClause}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  \Split{} and \Propagate{} cannot be applied \\
  \bottomrule
  \end{tabular}
  \caption{Derivation rules for one automaton.}\label{tbl:rules:single}
\end{table}

We use the symbolic function $\FlowEq(\Automaton, \Filter)$ that generates a set
of existentially quantified linear inequalities with the following definition:
$$
\begin{aligned}
  & \FlowEq(\Automaton, \Filter) = \sum\limits_{\State \in \AcceptingStates} \FinalStateVar_\State = 1 \land \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)} = \Sink(\State) \\
  & \Sink(\State) = 0 \text{ if } \State \not\in \AcceptingStates, \FinalStateVar_\State \text{ otherwise.} \\
  & \In(\State, \Filter) = \StartFlow(\State) + \sum_{\Transition = \FromLabelTo{\State_0}{}{\State} \in \Transitions} \Filter(\Transition)\\
  & \StartFlow(\State)  = 1 \text{ if } \State = \InitialState, \text{ otherwise $0$.} \\
  & \Out(\State, \Filter) = \sum_{\Transition = \FromLabelTo{\State}{}{\State'} \in \Transitions} \Filter(\Transition)
\end{aligned}
$$
Where we transparently assign fresh variables to each state $\State \in
\States$ as well as every accepting state $\State' \in \AcceptingStates$ and
existentially quantify them in the whole clause.

Additionally, we assume the existence of a rule \PresburgerClose{},
corresponding to a sound and complete solver for Presburger formulae, modulo the
monoid~$\Monoid$.

@@

A decision procedure for our predicate in a tableau-based automated theorem
prover would start by expanding the predicate using the $\Expand{}$~rule. A
modestly clever theorem prover would perform algebraic substitution on the
underlying constants of~$\Filter$, boiling them down to choices of branches,
which depend on one single variable, and loop transitions. This logic
corresponds to the placement of counters for optimally edge-profiling the CFG of
a program, making up a MST of the automaton~\cite{path-profiling}.

@@

\begin{figure}[ht]
  We define $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$ as follows:

  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{S}{a}{A}) & = \TransitionVar_1 \\
        & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_2 \\
        & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{B}{b}{S}) & = \TransitionVar_4 \\
        & \Filter(\FromLabelTo{A}{a}{F}) & = \TransitionVar_5 \\
        & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        % S
        &  \TransitionVar_4 = \TransitionVar_5 + \TransitionVar_2 - 1 \\
        % A
        & \TransitionVar_1 = \TransitionVar_5 \\
        % B
        & \TransitionVar_2 = \TransitionVar_4 + \TransitionVar_6 \\
        % F
        & \TransitionVar_5 + \TransitionVar_6 = \FinalStateVar_1 \\
        & \FinalStateVar_1 = 1 \\
      \end{aligned}  
    \end{equation*}    
  \end{minipage}
  \caption{Equivalences defining $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$
  respectively.}\label{fig:example:single:equivalences}
  \end{figure}

Propagating the equivalences from \FlowEq{} gives us (under the implicit
assumption that every RHS is $\geq 0$):
\begin{equation*}
  \begin{aligned}
    & \Filter(\FromLabelTo{S}{a}{A}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_4 + \TransitionVar_6 \\
    & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
    & \Filter(\FromLabelTo{B}{b}{S}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
    & \Filter(\FromLabelTo{A}{a}{F}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
  \end{aligned}
\end{equation*}

\begin{figure}
  \centering
\begin{prooftree}
  \infer0[\PresburgerClose]{
    \begin{matrix}
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Propagate]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 \geq 0 \land \\
      2 + \TransitionVar_4 \geq 0 \land \\
      2k = 2 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[Eq. reason]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix}
  }
  % BRANCH:  = 0
  \infer0[\PresburgerClose{}]{
    \begin{matrix}
      \TransitionVar_3 > 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_3 > = 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[Eq. reason]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0\\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Propagate{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[Eq. reason]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix} % BRANCH: > 0
  } % SPLIT
  \infer2[\Split{} $1 - \TransitionVar_6$]{ \Connected(\AcaOrBc{}, \Filter) \land 2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6 }
  \infer1[\Expand{}]{\Image{}_{\AcaOrBc{}, \Transition \mapsto 1}(\Filter, 2k + 1)}
\end{prooftree}
\caption{A derivation for \Calculus{} computing odd lengths in $\AcaOrBc{}$.}\label{fig:derivation:single}
\end{figure}

@@

To be as clear as possible about what it means for our calculus to be correct,
we use more exact definitions than the traditional soundness and completeness.
Since correctness is guaranteed by the formulation of the predicate, our
correctness proof means proving inductively that no rule would invalidate that
definition.

\subsubsection{\Calculus{} terminates}
\begin{lemma}\label{lma:single-terminates}
  A solver implementing the calculus described in \cref{sec:calculus} starting
  with some decidable (modulo the solver) set of clauses $\SomeClause{}$ and an
  instance of our predicate $\SinglePredicateInstance$ will eventually
  terminate.
\end{lemma}

>>>>

\subsection{Example~1: String length}
\Fudge{We have already illustrated refer intro}
A useful example of such a simplifying morphism can express string length, the
problem that originally motivated our study of the Parikh map. We let $\Monoid =
\left(\mathbb{Z};+;0\right)$ and let $\Map(x) = 1$. Then the length of a string
$s = s_1 \RepeatSum{\Concat}  s_n$ is given by $\Map(s) = \sum \Map(s_i) = 1
\RepeatSum{+} 1$. Note that if we substitute vectors and vector addition for
this monoid, we obtain the regular Parikh map.

Intuitively, this often produces a simpler problem than computing the Parikh
image, as we now ignore not just the order of characters but also their
differences.

\subsection{Example~2: Parikh Automata}\label{sec:parikh-automata}
\contents{\item Ostrich+!}

@@

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where $C \subseteq \Naturals^d$ is
  a semi-linear set (or, equivalently, a Presburger formula), and $\Automaton$
  is a finite automaton with the alphabet $\Alphabet \times D$, where $D
  \subseteq \Naturals^d$. We say that $\Tuple{\Automaton, \varphi}$
  \emph{recognises} the language, $\Language(\Automaton, C)$, that is it
  recognises the language created by is the $\pi_\Alphabet$-projection on
  $\Alphabet$ whose $\pi_{v} \subseteq C$.
\end{definition}

\begin{figure}[t]
  \centering
      \includegraphics[width=0.5\linewidth]{parikh_automaton}
      \caption{The automaton part of a Parikh automaton for $\AcaOrBc{}$ with
      $\Alphabet = \Set{\text{a, b, c}}, d = 3$. The semilinear set/Presburger
      formula containing the constraints on the final register values cannot be
      visualised.}\label{fig:parikh-automaton}
    \end{figure}

@@

We start by defining our calculus, \Calculus{}, for one automaton, and only
extend it to products of automata in \cref{sec:multiple}. Assume an NFA
$\Automaton = \AutomatonTuple$ with $\NrTransitions$ transitions $\Transitions =
\Set{\EllipsisSequence{\Transition}{\NrTransitions}}$. For convenience, we then
introduce the following supporting notations:

\begin{definition}
  A \textit{path} $\Path = \PathEnumeration$ of an automaton~$\Automaton$ with
  transitions $\Transitions$ represents a path through $\Automaton$ using
  transitions in $\Transitions$ (i.e. $\Transitions(\State_k, \Label_{k+1},
  \State_{k+1})$ holds), passing zero or more labels $\Label_1, \ldots,
  \Label_n$. The path must begin in the initial state, i.e.~$\State_0 =
  \InitialState$. The end state, $\State_n$, is not necessarily accepting. If it
  is, we say the path is also \textit{accepting}.
  \end{definition}

\begin{definition}
  Moreover, we talk about the \textit{set of paths} of an automaton,
  $\Paths(\Automaton)$, a possibly infinite (if $\Automaton$ has loops) set of
  valid paths through $\Automaton$. Additionally, we use the
  notation~$\Paths(\Automaton, \State)$ to mean all paths ending in
  state~$\State$.
\end{definition}

\begin{definition}
  For a path $\Path = \Tuple{\State_0 \Label_1 \State_1 \RepeatSum{,}
  \State_n}$, its \textit{word} $\WordOf(\Path) = \Label_1 \RepeatSum{\Concat}
  \Label_n$ is the word read out on its labels.
\end{definition}

\begin{definition}
  The \textit{states} of a path $\Path = \Tuple{\State_0 \Label_1 \State_1
  \RepeatSum{,} \State_n}$, $\StatesOf(\Path) = \Set{\State_0\RepeatSum{,}
  \State_n}$ are the states visited along $\Path$. Note that $\InitialState \in
  \StatesOf(\Path)$ for every path since all paths start in the initial state.
\end{definition}

\begin{definition}
  A \textit{cut}, $C$ of an automaton~$\Automaton = \AutomatonTuple$ to state
  $\State$, written $\SeparatingCut(C, \Automaton, \State)$, is a minimal set of
  transitions such that every accepting path $p$ where $\State \in
  \StatesOf(\Path)$ contains a transition $\Transition \in C$. A cut does not
  exist for every state; notably it never exists for $\InitialState$.
\end{definition}

@@

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. Note that we operate on sets of symbols (terms,
clauses). Additionally, we also use the convention of splitting the formulas
into linear inequalities ($\SomeInequalities$) and any other clauses
($\SomeClause$).

Note that the filtering function~$\Filter$ is evaluated symbolically in these
rules, and can in practice be read as a symbolic function from transitions to
terms (e.g.~\texttt{t} or~\texttt{t+1}). In our implemendation \Catra{},
described in \cref{sec:implementation}, $\Filter$~is a vector of fresh variables
with the same size as~$\Transitions$.

We use the shorthand notation~$\Transitions_\Automaton$ to refer to the
transitions of an automaton~$\Automaton$. Additionally, for an
automaton~$\Automaton = \AutomatonTuple$ we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$. Similarly, for our commutative monoid $\Monoid =
\left(X;+_{\Monoid};0_{\Monoid}\right)$ and the map into it $\Map : \Strings
\rightarrow X$, we also allow mapping over transitions:
$\Map(\FromLabelTo{\State}{\Label}{\State'}) = \Map(\Label)$.

The rules of \cref{tbl:rules:single} are meant to be used in tableaux-style
proofs, and are read bottom-up, where the syntax at the bottom is replaced with
the one at the top until \Fudge{it is possible to close the proof goal}. Since
our rules operate by adding and matching linear (in)equalities in a proof goal,
we use the shorthand of listing the matched inequalities as antecedents. An
example of this can be seen in the~\Propagate{} rule. \Fudge{We pull solutions
out of a hat? I have no idea.}

We require that every rule only fires when it would add a new clause. For
example, this means that we cannot use~\Split{} to split on the same term twice.
This suggests a proof strategy where you \Propagate{} when you can, \Split{}
when you must, and \Subsume{} when neither is possible anymore.

\begin{table}
\begin{tabular}{@{}l>{$}c<{$}p{3cm}@{}}\toprule
  Name & Rule & Prerequisites\\
  \midrule

  % EXPAND
  \Expand & 
    \inferrule
  {\Connected(\Automaton, \Filter) \land \FlowEq(\Automaton, \Filter) \land \MonoidElement = \sum_{\Transition \in \Transitions_\Automaton} \Filter(\Transition) \cdot \Map(\Transition), \SomeInequalities, \SomeClause}
  {\SinglePredicateInstance, \SomeInequalities, \SomeClause} & 
  None \\[3ex]

  % SPLIT
  \Split & 
  \inferrule{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) = 0 \mid \Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause, \Filter(\Transition) > 0}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  if $\Transition \in \Transitions_\Automaton$ \\[3ex]

  % PROPAGATE
  \Propagate &
  \inferrule{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause, \Filter(\Transition) = 0}{\Connected(\Automaton, \Filter), \Set{\Filter(\Transition') = 0 \SuchThat \Transition' \in C}, \SomeInequalities, \SomeClause} &
  if $t = \FromLabelTo{\State}{}{\State'} \in \Transitions_\Automaton, \SeparatingCut(C, \Automaton, \State)$ \\[3ex]

  % SUBSUME
  \Subsume &
  \inferrule{\SomeInequalities,\SomeClause}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  \Split{} and \Propagate{} cannot be applied \\
  \bottomrule
  \end{tabular}
  \caption{Derivation rules for one automaton.}\label{tbl:rules:single}
\end{table}
We use the symbolic function $\FlowEq(\Automaton, \Filter)$ that generates a set
of existentially quantified linear inequalities with the following definition,
where we assign fresh, existentially quantified variables to
$\FinalStateVar_\State, \Filter(\Transition)$ for every $\State \in
\AcceptingStates, \Transition \in \Transitions$:
\[
\begin{aligned}
  & \FlowEq(\Automaton, \Filter) = \sum\limits_{\State \in \AcceptingStates} \FinalStateVar_\State = 1 \land \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)} = \Sink(\State) \land
  \AndComp{\State \in \AcceptingStates}{\FinalStateVar_\State \geq 0}\\
  & \Sink(\State) = 0 \text{ if } \State \not\in \AcceptingStates, \FinalStateVar_\State \text{ otherwise.} \\
  & \In(\State, \Filter) = \StartFlow(\State) + \sum_{\Transition \in \FromLabelTo{\State'}{}{\State}} \Filter(\Transition)\\
  & \StartFlow(\State)  = 1 \text{ if } \State = \InitialState, \text{ otherwise $0$.} \\
  & \Out(\State, \Filter) = \sum_{\Transition \in \FromLabelTo{\State}{}{\State'}} \Filter(\Transition)
\end{aligned}
\]

In addition to \cref{tbl:rules:single}, we assume the existence of a rule
\PresburgerClose{}, corresponding to a sound and complete solver for Presburger
formulae, and for the elements of~$\Monoid$.

@@

A decision procedure for our predicate in a tableau-based automated theorem
prover would start by expanding the predicate using the $\Expand{}$~rule. A
theorem prover would perform algebraic substitution on the underlying constants
of~$\Filter$, boiling them down to choices of branches, which depend on one
single variable, and loop transitions. This logic corresponds to the placement
of counters for optimally edge-profiling the CFG of a program, making up a
minimum-spanning tree of the automaton~\cite{path-profiling}.

In order to make the examples below tractable, we will assume the existence of a
rule \EquationReasoning{} that allows us to perform standard algebraic reasoning
on linear inequalities. This rule is not necessary for correctness or
completeness, but shortens the examples considerably.

@@

\begin{figure}[ht]
  We define $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$ as in the following two
  equations, and then apply \EquationReasoning{} (under the implicit assumption
  that every RHS is $\geq 0$) to obtain the equivalent third definition :
  \begin{minipage}[b]{0.3\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{S}{a}{A}) & = \TransitionVar_1 \\
        & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_2 \\
        & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{B}{b}{S}) & = \TransitionVar_4 \\
        & \Filter(\FromLabelTo{A}{a}{F}) & = \TransitionVar_5 \\
        & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.3\linewidth}
    \begin{equation*}
      \begin{aligned}
        % S
        &  \TransitionVar_4 = \TransitionVar_5 + \TransitionVar_2 - 1 \\
        % A
        & \TransitionVar_1 = \TransitionVar_5 \\
        % B
        & \TransitionVar_2 = \TransitionVar_4 + \TransitionVar_6 \\
        % F
        & \TransitionVar_5 + \TransitionVar_6 = \FinalStateVar_1 \\
        & \FinalStateVar_1 = 1 \\
      \end{aligned}  
    \end{equation*}    
  \end{minipage}
  \begin{minipage}[b]{0.3\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{S}{a}{A}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_4 + \TransitionVar_6 \\
        & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{B}{b}{S}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        & \Filter(\FromLabelTo{A}{a}{F}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \caption{Equivalences defining $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$
  respectively.}\label{fig:example:single:equivalences}
  \end{figure}

\begin{figure}
  \centering
\begin{prooftree}
  \infer0[\PresburgerClose]{
    \begin{matrix}
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Propagate]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 \geq 0 \land \\
      2 + \TransitionVar_4 \geq 0 \land \\
      2k = 2 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix}
  }
  % BRANCH:  = 0
  \infer0[\PresburgerClose{}]{
    \begin{matrix}
      \TransitionVar_3 > 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_3 > = 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0\\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Propagate{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix} % BRANCH: > 0
  } % SPLIT
  \infer2[\Split{} $1 - \TransitionVar_6$]{ \Connected(\AcaOrBc{}, \Filter) \land 2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6 }
  \infer1[\EquationReasoning{}]{
    \Connected(\AcaOrBc{}, \Filter) \land
    2k + 1 =
    \sum\limits_{\Transition \in \Transitions} (\Transition \mapsto 1)(\Label) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{\AcaOrBc{}, \Transition \mapsto 1}(\Filter, 2k + 1)}
\end{prooftree}
\caption{A derivation for \Calculus{} computing odd lengths in $\AcaOrBc{}$.}\label{fig:derivation:single}
\end{figure}

@@

Since correctness is defined by the formulation of the $\Image{}$~predicate, our
correctness proof amounts to proving inductively that no rule would invalidate
that definition.

\subsubsection{\Calculus{} terminates}
\begin{lemma}\label{lma:single-terminates} A solver implementing the calculus
  described in \cref{sec:calculus} starting with some decidable set of clauses
  $\SomeClause{}$ and an instance of our predicate $\SinglePredicateInstance$
  will eventually terminate.
\end{lemma}
